**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-06 Schloss_Cluster_PeerJ_2015.Rmd
Add results for MiSeq data

## To Do:
* miseq analysis
* methods section
* references
* figure captions

@@

***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTUs may be meaningless by poorly representing a specified cutoff in the assignment of sequences to those OTUs. To assess the accuracy and precision of the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth so that we could calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 3A). The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms. Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL. Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same algorithm, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs. Our analysis indicates that the best algorithm for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL algorithm.

@@

n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences
```

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms. First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permits the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms. Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene. With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS). The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS). Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points.

[   Something, something, something.    ]

* Compare MCC and its variation when using different algorithms
* Address stability
* Quality of algorithms are data-dependent

@@

***Software availability***
QIIME v.1.9.1
mothur v.1.37.0
R v.3.2
Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015

>>>>

## To Do:
* Figure 5
* methods section
* references
* figure captions

@@

***Methods vary in their accuracy.*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTUs may be meaningless by poorly representing a specified cutoff in the assignment of sequences to those OTUs. To assess the accuracy and precision of the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth so that we could calculate the average MCC value as a measure of a method's accuracy and its variation as a measure of its precision. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL algorithms were the most sensitive; however, the MCC values using 80 and 100% of the data did not vary meaningfully when using *de novo* algorithms. Second, the AL algorithm out-performed the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL algorithm, the MCC values for each of the only demonstrated a small amount of variation between runs of the algorithm with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same algorithm, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the the full dataset is not optimally assigned to OTUs. Our analysis indicates that the best algorithm for assigning the Canadian soils sequences to OTUs using a 97% threshold is the AL algorithm.

@@

n_groups <- ncol(read.table(file="data/miseq/miseq.trim.contigs.good.unique.good.filter.unique.precluster.denovo.uchime.pick.pick.count_table")) - 2 #subtract 2 to remove the sequence name and total number of sequences

miseq_stability <- read.table(file="data/process/miseq.mcc_ref.summary", header=T)
miseq_mcc <- read.table(file="data/process/miseq.mcc.summary", header=T)
```

***Deep sampling of 16S rRNA genes.*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering algorithms. First, the Canadian soil dataset that the He study used to analyze the stability of OTUs is one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permits the sequencing of millions of sequences for a study. In addition, because the original phred quality scores and flowgram data are not available, it was not possible to adequately remove sequencing errors (REFS). The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering algorithms. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference algorithms. Of the `r nrow(gg_summary)` sequences in the default QIIME database, only `r n_v4_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene. With these points in mind, we compared the clustering assignment for each of these algorithms using a time series experiment that was obtained using mouse stool (REFS). The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% (REFS). Although no dataset is perfect for exhaustively testing these clustering algorithms, this dataset was useful for demonstrating several points. First, when using 60% of the data the stability relationships amongst the different methods were similar to what we observed using the He dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the He dataset for all of the methods except for CL and SL. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) algorithms were comparable to the AL algorithm (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal algorithm may be database-dependent. Finally, as was observed with the He dataset, there was little variation in MCC values observed among the 30 randomizations. Therefore, although the algorithms have a stochastic component, the OTU assignments do not vary meaningfully. The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* algorithms can generate very stable OTU assignments, as defined by the He study, and that the assignments are highly reproducible. Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* algorithms are consistently more robust than either of the reference-based methods.

@@

***Software availability***
QIIME v.1.9.1
mothur v.1.37.0 - includes sens.spec command and vsearch enabled clustering
R v.3.2
vsearch
Reproducible workflow including this manuscript as a literate programming document is available at https://github.com/SchlossLab/Schloss_Cluster_PeerJ_2015
